---
title: "Reproduceable Research Week 2 Assignment"
author: "Anas Mubashir"
date: "June 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data

### The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

* date: The date on which the measurement was taken in YYYY-MM-DD format

* interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

# Loading and preprocessing the data

### Loading the Data and processing it if necessary

```{r}
activity_data <- read.csv("activity.csv")
summary(activity_data)
head(activity_data)
```

```{r}
steps_by_date <- aggregate(activity_data$steps, by = list(Date = activity_data$date), FUN = sum)
names(steps_by_date)[names(steps_by_date) == "x"] <- "Steps"
temp_d <- as.Date(steps_by_date$Date, "%Y-%m-%d")
steps_by_date$Date <- format(temp_d, format = "%m-%d")
head(steps_by_date)
```

# What is mean total number of steps taken per day?

### Make a histogram of the total number of steps taken each day

```{r}
hist(steps_by_date$Steps, xlab = "Number of Steps Per Day", ylab = "Days", main = "Total Number of Steps taken per Day", col = "aquamarine")
```
### Calculate and report the mean and median total number of steps taken per day
```{r}
mean(na.omit(steps_by_date$Steps))
median(na.omit(steps_by_date$Steps))
```

# What is the average daily activity pattern?
### Average Daily Activity Pattern
```{r}
library(ggplot2)
steps_by_interval <- aggregate(steps ~ interval, data = activity_data, FUN =mean)
TimeSeries1 <- ggplot(data = steps_by_interval, aes(x = interval, y = steps)) + 
  geom_line() +
  xlab("Time Intervals") + 
  ylab("Total Number of Steps") +
  ggtitle("Average Number of Steps Taken per 5-Minute Interval")
print(TimeSeries1)
```

### The 5-minute interval that, on average, contains the maximum number of steps
```{r}
steps_by_interval[which(steps_by_interval$steps == max(steps_by_interval$steps)),]
```

# Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.
### Calculate and report the total number of missing values in the dataset
```{r}
missingVals <- sum(is.na(activity_data))
```

### Filling NAs with mean value.
```{r}
library(dplyr)
replace_with_mean <- function(num) replace(num, is.na(num), mean(num, na.rm = TRUE))
temp_data <- (activity_data %>% group_by(interval) %>% mutate(steps = replace_with_mean(steps)))
head(temp_data)
sum(is.na(temp_data))
```

### New dataset with the missing data filled in.
```{r}
new_activity_data <- as.data.frame(temp_data)
head(new_activity_data)
summary(new_activity_data)
```

### Making Histogram with new data set.
```{r}
new_steps_by_date <- aggregate(new_activity_data$steps, by = list(Date = new_activity_data$date), FUN = sum)
names(new_steps_by_date)[names(new_steps_by_date) == "x"] <- "Steps"
temp_d <- as.Date(new_steps_by_date$Date, "%Y-%m-%d")
new_steps_by_date$Date <- format(temp_d, format = "%m-%d")
head(new_steps_by_date)
hist(new_steps_by_date$Steps, xlab = "Number of Steps Per Day", ylab = "Days", main = "Total Number of Steps taken per Day", col = "red")
```

### Comparing mean median of both dataset.
```{r}
median(na.omit(steps_by_date$Steps))
median(na.omit(new_steps_by_date$Steps))
mean(na.omit(steps_by_date$Steps))
mean(na.omit(new_steps_by_date$Steps))
```

# Are there differences in activity patterns between weekdays and weekends?
### Create a new factor variable in the dataset.
```{r}
new_activity_data$WeekendOrWeekday <- ifelse(weekdays(as.Date(new_activity_data$date)) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "Weekday", "Weekend")
head(new_activity_data)
```

### Creating a panel plot
```{r}
new_activity_data <- (new_activity_data %>% group_by(interval, WeekendOrWeekday) %>% summarise(Mean = mean(steps)))
ggplot(new_activity_data, mapping = aes(x = interval, y = Mean)) + geom_line() +
  facet_grid(WeekendOrWeekday ~.) + xlab("Interval") + ylab("Mean of Steps") +
  ggtitle("Comparison of Average Number of Steps in Each Interval")
```

